---
title: "EDA-challange"
format: html
editor: visual
---

**STEP1**

```{r}
library(tidyverse)
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/main/data-wrangling.csv"
d <- read_csv(f, col_names = TRUE )

colnames(d)
glimpse(d)
```

**1,2,3**

```{r}
# 1. Create BSD 
BSD <- d$Body_mass_male_mean / d$Body_mass_female_mean

# 2. Create Sex Ratio 
sex_ratio <- d$AdultFemale / d$AdultMales

# 3. Create Defensibility Index 
DI <- d$DayLength_km / (sqrt(d$HomeRange_km2 / pi))

```

***4***

```{r}
# Overall
library(ggplot2)
library(dbplyr)

p <- ggplot(data = d, aes(x = log(Move), y = log(DayLength_km),
    color = "orange"))  
p <- p + xlab("log(Move)") + ylab("log(DayLength_km)")
p <- p + geom_point(na.rm = TRUE)
p <- p + theme(legend.position = "bottom", legend.title = element_blank())
p <- p + geom_smooth(method = "lm", fullrange =FALSE, na.rm =TRUE)
p


# By family
p <- p + facet_wrap(~Family)
     p + theme(legend.position = "none")

```

***Do species that spend more time moving travel farther overall?***

Looking at the overall trend across all families,species that live in larger groups travel farther.

***How about within any particular primate family?***

There are positive correlation in & species. Hylobatidae and Atilidae have positive correlation.

***Should you transform either of these variables?***

Yes, transformated values make clear results for the population distrubition

***5***

```{r}
#Overall

library(ggplot2)
p <- ggplot(data = d, aes(x = log(MeanGroupSize),
                          y =    log(DayLength_km), 
                          color = "blue"))
     
p <- p + xlab("MeanGroupSIze)") + ylab("DayLength_km)")
p <- p + geom_point(na.rm = TRUE)
p <- p + theme(legend.position = "bottom", legend.title = element_blank())
p <- p + geom_smooth(method = "lm", fullrange =FALSE, na.rm =TRUE)
p

#by family
p <- p + facet_wrap(~Family)
     p + theme(legend.position = "none")
```

***Do species in larger groups travel farther overall?***
Looking at the overall trend across all families,species that live in larger groups travel farther.
 
***How about within specific primate families?***

In Cercopithecidae, there's a noticeable increase in day range length with group size.

***Should we transform either variable?***

Yes, transformated values make clear results for the population distrubition



***6*** 
```{r}
##Overall

library(ggplot2)
p <- ggplot(data = d, aes(x = log(Canine_Dimorphism),
                          y =    log(BSD), 
                          color = "blue"))
     
p <- p + xlab("Body Size Dimorphism") + ylab("Canine_Dimorphism")
p <- p + geom_point(na.rm = TRUE)
p <- p + theme(legend.position = "bottom", legend.title = element_blank())
p <- p + geom_smooth(method = "lm", fullrange =FALSE, na.rm =TRUE)
p

#by family
p <- p + facet_wrap(~Family)
     p + theme(legend.position = "none")
```

**Do taxa with greater size dimorphism also show greater canine dimorphism?**

Yes. Taxa with greater size dimorphism shows greater canine dimorphism.


**7**

```{r}
d <- mutate(d, diet_strategy = ifelse(Fruit >= 50, "frugivore", 
                                      ifelse(Leaves >= 50, "folivore", 
                                             ifelse(Fruit < 50 & Leaves < 50, "omnivore", NA))))


glimpse(d)

# Cleaning NA
d_clean <- d[complete.cases(d[, c("diet_strategy", "MeanGroupSize")]), ]
glimpse(d_clean)

# Boxplot
p <- ggplot(d_clean, aes(x = diet_strategy, y = log(MeanGroupSize))) +
  geom_boxplot() +
  geom_jitter(color = "blue", width = 0.1) + 
  labs(x = "Diet Strategy", y = "Log of Mean Group Size") +
  theme(axis.text.x = element_text(angle = 90))

p
```

**Do frugivores live in larger groups than folivores?**

No. Frugivores live in smaller groups than foliveres.

***8***
```{r}

s <- mutate(d, Binomial = paste(Genus, Species, sep = " ")) %>%
     select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) %>%
     group_by(Family) %>%
     summarise(
     avg_BrainSize = mean(Brain_Size_Species_Mean, na.rm = TRUE),
     avg_BodyMass = mean(Body_mass_male_mean, na.rm = TRUE)) %>%
     arrange(avg_BrainSize)

s
```
